---
title: "Hungarian lakok/lakom: a worked example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hungarian lakok/lakom: a worked example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Background

Hungarian verbs ending in a labial consonant alternate between two first-person
singular forms: *lakok* and *lakom*. Speakers show gradient preferences — some
verbs strongly favour one form, others are variable. This vignette uses kernel
ridge regression (KRR) on phonological similarity to predict those preferences
for both attested corpus verbs and novel nonwords.

Data come from [Rácz (2024)](https://github.com/petyaracz/Racz2024). See
`system.file("scripts/setup.R", package = "krr")` for how the training data
were derived from raw corpus counts and phonological distances computed via
[JANET](https://github.com/petyaracz/JANET).

## Load

```{r load, message = FALSE}
library(krr)
library(readr)
library(dplyr)
library(ggplot2)
```

## Data

The package ships with three files in `inst/extdata/`:

```{r data, message = FALSE}
train <- read_tsv(system.file("extdata", "lakok_train.tsv",       package = "krr"))
test  <- read_tsv(system.file("extdata", "lakok_test.tsv",        package = "krr"))
dist  <- read_tsv(system.file("extdata", "word_distances.tsv.gz", package = "krr"))
```

`train` — corpus verbs with columns `lemma` (phonological form), `base`
(orthographic lemma), `lemma_freq_corrected` (corpus frequency), and `p`
(probability of *lakok*).

`test` — nonword experimental ratings with `lemma` and `p`.

`dist` — pairwise aligned phonological distances (`word1`, `word2`, `phon_dist`)
for all lemmas in both datasets.

## Prepare training data

Each base lemma may appear in multiple inflected corpus forms. Keep the
highest-frequency form per base so each base contributes exactly one training
observation:

```{r dedup}
train <- train |>
  group_by(base) |>
  arrange(-lemma_freq_corrected) |>
  slice(1) |>
  ungroup()
```

## Check inputs

`check_krr_inputs()` catches common failure modes — duplicate words, missing
distance pairs, outcome values incompatible with the link — before they produce
opaque matrix algebra errors:

```{r check}
check_krr_inputs(train, test, dist, word_col = "lemma", outcome_col = "p", link = "logit")
```

## Train

`train_krr()` performs a grid search over the RBF bandwidth (`sigma`) and ridge
penalty (`alpha`) using leave-one-out cross-validation. LOO predictions are
computed efficiently via the hat-matrix shortcut (PRESS formula), so no refitting
is needed per fold:

```{r train, cache = TRUE}
m <- train_krr(
  train, dist,
  word_col     = "lemma",
  outcome_col  = "p",
  link         = "logit",
  criterion    = "rmse"
)
cat("Best sigma:", m$sigma, " Best alpha:", m$alpha, "\n")
cat("LOO RMSE (logit space):", round(m$best_score, 4), "\n")
```

## LOO fit on training data

```{r loo-plot, fig.width = 5, fig.height = 4}
m$predictions |>
  ggplot(aes(observed, predicted_loo)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x     = "Observed p(lakok)",
    y     = "LOO-predicted p(lakok)",
    title = "Training data: LOO fit"
  )
```

## Predict nonwords

```{r predict}
preds <- predict_krr(
  train, test, dist,
  word_col    = "lemma",
  outcome_col = "p",
  sigma       = m$sigma,
  alpha       = m$alpha,
  link        = "logit"
)
```

```{r pred-plot, fig.width = 5, fig.height = 4}
preds |>
  ggplot(aes(observed, predicted)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x     = "Observed p(lakok) — nonword ratings",
    y     = "Predicted p(lakok)",
    title = "Nonword predictions"
  )
```

```{r cor}
with(preds, cor.test(observed, predicted))
```

If you can reproduce these results you are ready to apply `krr` to your own
data. See `?train_krr` and `?predict_krr` for the full API, and the `ex/`
directory in the package source for examples from other languages.
